Commutative Character Sheaves

### 2015-09-19 followup.tex
Commit after talking with David.
Moved material on the derived group to section defining CCS

@@

An $H$-equivariant rank-$1$ character sheaf on $H$ is a rank-$1$ character sheaf $\cs{L} = (\gcs{L},\phi,\mu)$ on $G$ together with an isomorphism  $\nu : {\bar m}_H^*\gcs{L} \to {\bar p}_H^*\gcs{L}$ such that the following diagrams commute, expressing compatibility with $\phi$ and $\mu$:
\begin{enumerate}
\item
\[
\begin{tikzcd}
\ &  m_0^*  m^*\, \mathcal{L} \arrow{r}{m_0^*\nu} & m_0^* p^*\, \mathcal{L} & \\
m_1^* m^*\, \mathcal{L} \arrow{dr}[swap]{m_1^*\nu} \arrow[equal]{ur} &&& \arrow[equal]{ul}  \arrow{dl}{p_2^*\nu} p_2^* m^*\, \mathcal{L} \\
& m_1^* p^*\, \mathcal{L} \arrow[equal]{r} & p_2^* p^*\, \mathcal{L} & 
\end{tikzcd}
\]
where $m_0(a,b,x) = (a,m(b,x))$ and $m_1(a,b,x) = (m(ab),x)$ and $p_2(a,b,x) = (b,x)$;
\item
\end{enumerate}
Write $\CCS_H(G)$ for the category of $H$-equivariant commutative character sheaves on $G$.
\subsection{Descent}

@@

\begin{proof}
By descent theory, the functor $\Loc(G') \to \Loc_H(G)$ induced by $q: G \to G'$ is an equivalence of categories. 
\end{proof}

\begin{proof}
The group homomorphism $\pi_1(f): \pi_1(X, {\bar x}) \to \pi_1(Y, {\bar y})$ is surjective, for all geometric points with $f({\bar x}) = {\bar y}$ ... 
%\todo{Under construction}
\end{proof}

\begin{lemma}\label{lem:Lder}
For every $\cs{L}$ on $G$, $\cs{L}\vert_{G\der} \iso (\EE)_{G\der}$.
\end{lemma}

\begin{proof}
By construction, $c : G \times G \to G\der$ is dominant. 
Note that $j\circ c = m\circ (m \times (\iota\circ m\op)) : G \times G \to G$ is $(g,h) \mapsto gh g^{-1}h^{-1}$ while $m\op\circ (m \times (\iota\circ m\op))= e : G \times G \to G$ is $(g,h) \mapsto 1_G$. Thus,
\begin{align*}
c^* (\cs{L}\vert_{G\der})
&=  (j\circ c)^*\cs{L} \\
&= (m\circ (m \times (\iota\circ m\op)))^* \cs{L} \\
&= (m \times (\iota\circ m\op))^* m^* \cs{L} \\
&\iso (m \times (\iota\circ m\op))^* m\op^* \cs{L} \\
&= (m\op\circ (m \times (\iota\circ m\op)))^* \cs{L} \\
&= e^* \cs{L}\\
&= (\EE)_{G\times G}.
\end{align*}
Since $c : G \times G \to G\der$ is schematically dense, $c^* (\cs{L}\vert_{G\der}) \iso (\EE)_{G\times G}$ implies $\cs{L}\vert_{G\der} \iso (\EE)_{G\der}$ by Lemma~\ref{lem:dominant}.
\end{proof}

@@

By definition, the abelianization $G\ab$ of $G$ is the fppf-quotient $q : G \to G\ab$ of $G$ by $G\der$. 
This exists in schemes since  . . . \todo{I hope smoothness does the trick here.}
%\todo{Proof needed}
%  is the coequalizer of $n : G\der \times G \to G$ with projection $p: G\der \times G \to G$. 
\[
\begin{tikzcd}
G\der\times G
 \arrow{r}[shift left]{n}
  \arrow{r}[shift right,swap]{p}
&
G 
\arrow{r}{q}
& 
G\ab
\end{tikzcd}
\]

@@

\begin{proof}
The quotient $G \to G\ab$ is a $G\der$-torsor in the etale topology, \todo{Either true or it becomes a condition on $G$. Again, I hope smoothness does the trick.} so the result follows from Lemma~\ref{lem:descent}.
\end{proof}

\begin{theorem}\label{thm:Gab-descent}
 Let $G$ be a smooth group scheme over $\Fq$ 
 and let $q : G \to G\ab$ be its abelianization. 
 Then  
\[
q^*:  \CCS(G\ab) \longrightarrow \CCS(G)
\]
is an equivalence of rigid monoidal categories.
\end{theorem}
\begin{proof}
Combine Propositions~\ref{prop:Gder-equivariance} and \ref{prop:Gab-descent}.
\end{proof}

 \begin{corollary}
The following is a commuting diagram of isomorphisms:
\[
\begin{tikzcd}
\CCSiso{G} \arrow{r} & G(\Fq)^*\\
\arrow{u}{q^*} \CCSiso{G\ab} \arrow{r} & \arrow{u} G\ab(\Fq)^* . \\
\end{tikzcd}
\]
 \end{corollary}

 \begin{proof}
 By Theorem \ref{thm:trfrobiso}, $\CCSiso{G}\to G(\Fq)^*$ and $\CCSiso{G\ab}\to G\ab(\Fq)^*$ are isomorphisms.
By Theorem~\ref{thm:Gab-descent}, $\CCSiso{G} \iso \CCSiso{G\ab}$.
Since $G\der$ is simply connected, \todo{Proof needed} $G\ab(\Fq) = G(\Fq)/G\der(\Fq)$.
Thus, all characters of $G(\Fq)$ factor through
 $G\ab(\Fq)$. . . . 
 The isomorphisms are compatible because  . . . 
 \end{proof}

>>>>

\begin{proof}
The group homomorphism $\pi_1(f): \pi_1(X, {\bar x}) \to \pi_1(Y, {\bar y})$ is surjective, for all geometric points with $f({\bar x}) = {\bar y}$ ... 
\todo{Under construction}
\end{proof}

\begin{lemma}\label{lem:Lder}
If $\cs{L}\in \CCS(G)$ then $\cs{L}\vert_{G\der} \iso (\EE)_{G\der}$.
% Suppose Z(G) = 1$. If $\cs{L}\in \CS(G)$ and $\cs{L}\vert_{G\der} \iso (\EE)_{G\der}$ then $\cs{L}\in \CCS(G)$. 
\end{lemma}

\begin{proof}
By construction, $c : G \times G \to G\der$ is dominant. 
Note that $j\circ c = m\circ (m \times (\iota\circ m\op)) : G \times G \to G$ is $(g,h) \mapsto gh g^{-1}h^{-1}$ while $m\op\circ (m \times (\iota\circ m\op))= e : G \times G \to G$ is $(g,h) \mapsto 1_G$. 
Pick $\cs{L}\in \CCS(G)$. 
Then
\begin{align*}
c^* (\cs{L}\vert_{G\der})
&=  (j\circ c)^*\cs{L} \\
&= (m\circ (m \times (\iota\circ m\op)))^* \cs{L} \\
&= (m \times (\iota\circ m\op))^* m^* \cs{L} \\
&\iso (m \times (\iota\circ m\op))^* m\op^* \cs{L} \\
&= (m\op\circ (m \times (\iota\circ m\op)))^* \cs{L} \\
&= e^* \cs{L}\\
&= (\EE)_{G\times G}.
\end{align*}
Since $c : G \times G \to G\der$ is schematically dense, $c^* (\cs{L}\vert_{G\der}) \iso (\EE)_{G\times G}$ implies $\cs{L}\vert_{G\der} \iso (\EE)_{G\der}$ by Lemma~\ref{lem:dominant}.

%Now, suppose $Z(G) =1$. Pick $\cs{L}\in \CS(G)$ and suppose $\cs{L}\vert_{G\der} \iso (\EE)_{G\der}$.
\end{proof}

\begin{remark}
Lemma~\ref{lem:Lder} provides another indication of the correctness of Definition~\ref{def:CCS} 
if one is looking for the geometrization of characters of smooth group schemes over finite fields: commutative character sheaves are trivial on $G\der$, rank-$1$ character sheaves are not.
In the next section we provide examples to show that $\CCS(G)$ and $\CS(G)$ are not isomorphic categories, even when $G$ is a connected reductive algebraic group. 
\end{remark}

@@

An $H$-equivariant rank-$1$ character sheaf on $G$ is a rank-$1$ character sheaf $\cs{L} = (\gcs{L},\phi,\mu)$ on $G$ together with an isomorphism  $\nu : {\bar m}_H^*\gcs{L} \to {\bar p}_H^*\gcs{L}$ such that the following diagrams commute, expressing compatibility with $\phi$ and $\mu$:
\begin{enumerate}
\item
\[
\begin{tikzcd}
\ &  m_0^*  m^*\, \mathcal{L} \arrow{r}{m_0^*\nu} & m_0^* p^*\, \mathcal{L} & \\
m_1^* m^*\, \mathcal{L} \arrow{dr}[swap]{m_1^*\nu} \arrow[equal]{ur} &&& \arrow[equal]{ul}  \arrow{dl}{p_2^*\nu} p_2^* m^*\, \mathcal{L} \\
& m_1^* p^*\, \mathcal{L} \arrow[equal]{r} & p_2^* p^*\, \mathcal{L} & 
\end{tikzcd}
\]
where $m_0(a,b,x) = (a,m(b,x))$ and $m_1(a,b,x) = (m(ab),x)$ and $p_2(a,b,x) = (b,x)$;
\item
\end{enumerate}
Write $\CCS_H(G)$ for the category of $H$-equivariant commutative character sheaves on $G$.

\subsection{Descent}

@@

\begin{proof}
By descent theory, the functor $\Loc(G') \to \Loc_H(G)$ induced by $q: G \to G'$ is an equivalence of categories.
\todo{Need to study the extra structure coming with character sheaves.}
\end{proof}

@@

By definition, the abelianization $G\ab$ of $G$ is the fppf-quotient $q : G \to G\ab$ of $G$ by $G\der$. 
This exists in schemes and is a $G\der$-torsor in the etale topology since  . . . \todo{Either true or it becomes a condition on $G$. I hope smoothness does the trick. I've written to Masoud for help on this.}
%\todo{Proof needed}
%  is the coequalizer of $n : G\der \times G \to G$ with projection $p: G\der \times G \to G$. 
\[
\begin{tikzcd}
G\der\times G
 \arrow{r}[shift left]{n}
  \arrow{r}[shift right,swap]{p}
&
G 
\arrow{r}{q}
& 
G\ab
\end{tikzcd}
\]

@@

\begin{proof}
The quotient $G \to G\ab$ is a $G\der$-torsor in the etale topology,  so the result follows from Lemma~\ref{lem:descent}.
\end{proof}

\begin{theorem}\label{thm:Gab}
 Let $G$ be a smooth group scheme over $\Fq$ 
 and let $q : G \to G\ab$ be its abelianization. 
 Then  
\[
q^*:  \CCS(G\ab) \longrightarrow \CCS(G)
\]
is an equivalence of rigid monoidal categories.
\end{theorem}
\begin{proof}
Combine Propositions~\ref{prop:Gder-equivariance} and \ref{prop:Gab-descent}.
\end{proof}

 \begin{corollary}
The following diagram of isomorphisms commutes.
\[
\begin{tikzcd}
\CCSiso{G} \arrow{r} & G(\Fq)^*\\
\arrow{u}{q^*} \CCSiso{G\ab} \arrow{r} & \arrow{u} G\ab(\Fq)^* 
\end{tikzcd}
\]
 \end{corollary}

 \begin{proof}
 By Theorem \ref{thm:trfrobiso}, $\CCSiso{G}\to G(\Fq)^*$ and $\CCSiso{G\ab}\to G\ab(\Fq)^*$ are isomorphisms.
By Theorem~\ref{thm:Gab}, $\CCSiso{G} \iso \CCSiso{G\ab}$.
Since $G\der$ is simply connected, \todo{Proof needed} $G\ab(\Fq) = G(\Fq)/G\der(\Fq)$.
Thus, all characters of $G(\Fq)$ factor through
 $G\ab(\Fq)$. . . . 
 The isomorphisms are compatible because  . . . 
 \end{proof}

\begin{remark} 
It is not true that the categories $\CS(G\ab)$ and $\CS(G)$ are equivalent, even when $G$ is a connected reductive algebraic group, as we will see in the next subsection.
\end{remark}
